A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-03 paper/paper.tex
progress on syntax re-org

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{OHol}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\isSmall}{\,\,\textrm{Small}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

\newcommand{\punitinXfromYtoZ}[3]{#2 \mathop{\to_{#1}} #3}
\newcommand{\punitrefl}[1]{\textrm{id}_{#1}}
\newcommand{\punitelimtoYwithkontZ}[2]{\textrm{ind}_{\to}(#1,#2)}

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\to}(#4,#2,#1,#3,#5)}

@@

\begin{figure}
  TODO
  \caption{Set Connectives}
\end{figure}

@@

\begin{figure}

  \caption{Type Connectives}
\end{figure}

Finally, we describe the \emph{end}, $\pendallXdotY {\alpha} P$, which
when combined with the left or right hom types described in the next
section allows for us to internalize arbitrary transformation sequents
$\Phi \vdash P$ as a type.
%
This type is well formed when $\alpha$ is the covariant \emph{and}
contravariant variable of $P$.
%
An element of the type is a polymorphic element of $P$, varying in
$\alpha$.
%
The introduction form is naturally a $\lambda$ abstracting over
$\alpha:\cat C$ and the elimination form instantiates $\alpha$ with an
arbitrary object $a : \cat C$.
%
We also include the necessary $\beta\eta$ equalities.

\begin{figure}
  \begin{mathpar}
    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash P \isaSet}
    {\Gamma \vdash \pendallXdotY {\alpha:{\cat C}} P \isaTy}

@@

    \inferrule*[right=2Cell$\eta$]
    {\Gamma \vdash M : \pendallXdotY \alpha P }
    {\Gamma \vdash M = \pendlambdaXdotY \alpha {\pendappXtoY M \alpha} : \pendallXdotY \alpha P}
  \end{mathpar}
  \caption{End}
  \label{fig:end}
\end{figure}

Semantically, this type corresponds to taking the end of a profunctor
$P : {\cat C}^{op} \times {\cat C} \to \Set$, typically written
\[ \int^{\alpha:\cat C} P \alpha \alpha \]

@@

\subsection{Internalizing Virtual Equipment Structure}

\begin{figure}
  \begin{mathpar}
    \inferrule
    {\Gamma \vdash \cat C : \Sort \and\Gamma \vdash \cat D : \Sort}
    {\Gamma\pipe\cdot \vdash \varr{\cat C}{\cat D} \isaSet}

@@

    (\lambda\alpha:C. A) B = A[B/\alpha]

    \inferrule*[right=VArr $\eta$]
    {\Gamma \pipe\cdot \vdash M : \varr C D}
    {\Gamma \pipe\cdot \vdash M = (\lambda \alpha:C. M \alpha) : \varr C D}
    \\

@@

    (\lambda \alpha:C;\beta:D. R)A;B = R[A/\alpha;B/\beta]

    \inferrule*[right=HArr $\eta$]
    {\Gamma \pipe \cdot \vdash M : \harr C D}
    {\Gamma \pipe \cdot \vdash M = (\lambda \alpha;\beta. M \alpha;\beta) : \harr C D}\\

  \end{mathpar}
  \caption{Internalizing VE Structure}
\end{figure}

%% \subsection{Equalizers}

%% To encode the concept of internal hom of profunctors, kan extensions
%% and (co-)limits, we need to define the \emph{equalizer} of horizontal
%% arrows. Specifically, the universal property that we need is an
%% equalizer in the category whose objects are horizontal arrows (for
%% fixed endpoints) and whose arrows are the 2-cells.

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule*[right=Equalizer-formation]
%%     {\Gamma \pipe \cdot \vdash P : \harr\, C\,D\and
%%      \Gamma \pipe \cdot \vdash Q : \harr\, C \, D\\\\
%%      \Gamma \pipe \cdot \vdash \phi : \homunary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \alpha \beta}\and
%%      \Gamma \pipe \cdot \vdash \psi : \homunary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \alpha \beta}\and}
%%     {\Gamma \pipe \cdot \vdash \textrm{Equalizer}(C,D,P,\phi,\psi) : \harr\, C\, D}\and

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}
%%     {\Gamma \pipe \Phi \vdash \textrm{forget}(M) : \harrapp P A B}

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \harrapp P A B \and \appunary \phi A M B = \appunary \psi A M B}
%%     {\Gamma \pipe \Phi \vdash \textrm{equalizing}_{\phi,\psi}(M) : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}

%%     TODO: obvious equations
%%   \end{mathpar}
%%   \caption{Equalizers (Logical Framework style)}
%% \end{figure}

%% \begin{figure}
%%   \begin{mathpar}
%%     \inferrule*[right=Equalizer-formation]
%%     {\Gamma \pipe \Xi \vdash P \isaSet\and\Gamma \pipe \Xi \vdash Q \isaSet\and
%%       \Gamma\pipe d^-\Xi, p : P, d^+ \Xi \vdash \phi : Q\and
%%       \Gamma\pipe d^-\Xi, p : P, d^+ \Xi \vdash \psi : Q }
%%     {\Gamma \pipe \Xi \vdash \equalizeVofWbyXeqYatZ p P \phi \psi Q : \harr\, C\, D}\and

%%     \inferrule*[right=Equalizer-formation]
%%     {\Gamma \pipe \alpha:C;\beta:D \vdash P \isaSet\and\Gamma \pipe \alpha:C;\beta:D \vdash Q \isaSet\and
%%       \Gamma \pipe d^-\Xi \vdash A : C\and
%%       \Gamma \pipe d^+\Xi \vdash B : D\\\\
%%       \Gamma\pipe \cdot \vdash \phi : \homunary \alpha P \beta Q\and
%%       \Gamma\pipe \cdot \vdash \psi : \homunary \alpha P \beta Q
%%     }
%%     {\Gamma \pipe \Xi \vdash \equalizeVofWbyXeqYatZ p {P[A/\alpha;B/\beta]} {\appunary \phi A p B} {\appunary \psi A p B} {Q[A/\alpha;B/\beta]} : \harr\, C\, D}\and

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \equalizeVofWbyXeqYatZ p P \phi \psi Q}
%%     {\Gamma \pipe \Phi \vdash \pi_{eq}\,M : P}

%%     \inferrule
%%     {\Gamma \pipe \Phi \vdash M : \harrapp P A B \and \appunary \phi A M B = \appunary \psi A M B}
%%     {\Gamma \pipe \Phi \vdash \textrm{equalizing}_{\phi,\psi}(M) : \harrapp{\textrm{Equalizer}(C,D,P,\phi,\psi)} A B}

%%     TODO: equations
%%   \end{mathpar}
%%   \caption{Equalizers (Natural Deduction Style)}
%% \end{figure}

\subsection{Products}

We add in parallel two different notions of product: products of sorts
and products of spans, which have the universal property in their
respective categories.

\begin{figure}
  \begin{mathpar}
    \inferrule{}{\Gamma \vdash 1 : \Sort}\and
    \inferrule*{}{\Gamma\pipe\alpha:C \vdash () : 1}\and
    \inferrule*{\Gamma\pipe\alpha:C \vdash A : 1}{A = ()}
    \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}\and
    \inferrule{}{\Gamma\pipe \Phi \vdash () :  1}\and
    \inferrule{\Gamma\pipe \Phi \vdash M :  1}{M = ()}\\\\

    \inferrule{\Gamma \vdash C : \Sort \and \Gamma \vdash D : \Sort}
              {\Gamma \vdash C \times D : \Sort}\and

    \inferrule{\forall i \in \{1,2\}.~ \Gamma \pipe \alpha:C \vdash A_i : D_i}{\Gamma \pipe \alpha:C \vdash (A_1,A_2) : D_1 \times D_2}

    \inferrule{\Gamma \pipe \alpha:C \vdash A : D_1 \times D_2}{\Gamma \pipe \alpha:C \vdash \pi_i A : D_i}

    \pi_i(A_1,A_2) = A_i\\
    \inferrule{\Gamma\pipe \alpha:C \vdash A :  D_1\times D_2}{A = (\pi_1 A,\pi_2 A)}\\\\

    \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \and \Gamma\pipe\Xi \vdash S\isaSet}
              {\Gamma\pipe\Xi \vdash R \times S\isaSet}\and

    \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}

    \pi_i(M_1,M_2) = M_i\\
    \inferrule{\Gamma\pipe\Phi \vdash M :  R_1\times R_2}{M = (\pi_1 M,\pi_2 M)}
  \end{mathpar}
  \caption{Products}
  \label{fig:products}
\end{figure}

One additional connective we might desire is a ``parallel'' product of
sets $R \times' S : \harr {C_1 \times C_2} {D_1 \times D_2}$ where $R
: \harr {C_1} {D_1}$ and $R : \harr {C_2} {D_2}$. We don't need to
define this as a primitive though because it is expressible in terms
of the two kinds of products we have above:

\[
\inferrule
{\Gamma \pipe \alpha : C_i; \beta : D_i \vdash R_i \isaSet}
{\Gamma \pipe \alpha : C_1\times C_2; \beta : D_1 \times D_2
  R_1[\pi_1\alpha/\alpha;\pi_1\beta/\beta] \times R_2[\pi_2\alpha/\alpha;\pi_2\beta/\beta]\isaSet}
\]

The empty products give us access now to several traditional notions
of category theory that we couldn't before define, such as objects and
presheaves of either variance:

$$\textrm{Object}\,C = \varr 1 C$$
$$\textrm{Presheaf}\,C = \harr 1 C$$
$$\textrm{coPresheaf}\,C = \harr C 1$$

However, we find that in practice we don't have much use for these
concepts as most theorems about them easily generalize to more general
theorems about arbitrary functors or profunctors.

%% %% To define internal ends and coends, we need to go beyond the usual
%% %% simple types and add equalizers (a form of subsets) and co-equalizers
%% %% (a form of quotients) of judgments. Here's a sketch of the syntax for
%% %% each. These make type checking very difficult as they result in
%% %% obligations for proving equality of certain terms, which may or may
%% %% not be decidable.

%% %% \begin{mathpar}
%% %%   \inferrule*[right=Equalizer Formation]
%% %%   {\alpha;\beta\,|\, R \and
%% %%   x : R \vdash t : S x:R \vdash u : S}
%% %%   {\alpha;\beta\,|\,\{ x : R \,|\, t = u \}}

%% %%   \inferrule*[right=Equalizer Intro]
%% %%   {\Phi \vdash r : R \and t[r/x] = u[r/x]}
%% %%   {\Phi \vdash \textrm{assert}(x.t = u)(r) : \{ x : R \,|\, t = u \}}

%% %%   \inferrule*[right=Equalizer Elim]
%% %%   {\Phi \vdash r : \{ x : R \,|\, t = u \}}
%% %%   {\Phi \vdash \textrm{forget}(r) : R}

%% %%   t[\textrm{forget}(r)/x] = u[\textrm{forget(r)}/x]

%% %%   \textrm{assert}(x.t = u)(\textrm{forget}(r)) = r
%% %%   \textrm{forget}(\textrm{assert}(x.t = u)(r)) = r
%% %% \end{mathpar}

%% %% \begin{mathpar}
%% %%   \inferrule*[right=Co-equalizer Formation]
%% %%   {\alpha;\beta\,|\, R \and
%% %%   x : S \vdash t : R x:S \vdash u : R}
%% %%   {\alpha;\beta\,|\,R/(\forall x:S. t = u)}

%% %%   \inferrule*[right=Coequalizer Intro]
%% %%   {\Phi \vdash r : R}
%% %%   {\Phi \vdash \textrm{equate}(x:S.t=u)(r) : R/(\forall x.t=u)}

%% %%   \inferrule*[right=Coequalizer Elim]
%% %%   {\Psi \vdash r : R/(x.t = u) \and
%% %%     \Phi,y:R,\Phi' \vdash s : S \and
%% %%     s[t/y] = s[u/y]}
%% %%   {\Phi,\Psi,\Phi' \vdash \textrm{lett} \textrm{equate}(y) = r.~s : S}

%% %%   \textrm{equate}(x:S.t = u)(t[\gamma]) = \textrm{equate}(x:S.t = u)(u[\gamma])

%% %%   \cdots
%% %% \end{mathpar}

%% %% \subsection{Positive Multiplicatives}

%% %% \subsection{Negative Multiplicatives}

%% %% \subsection{Additives}

@@

One alternative to proarrow equipments is to axiomatize the bicategory
of categories and profunctors rather than the double category of
categories functors and profunctors. One abstract structure along
these lines is a \emph{Cartesian bicategory}\footnote{which is not the
same as a bicategory with products/finite limits, for which a syntax
was developed \citet{fioresaville}} is given in
\citet{carboni-walters,carboni-walters-someonelese}. Some preliminary
work on a proof system for Cartesian bicateogires was presented at
\citet{jonasfreycartesianbicategories}. This bears strong similarities
to our calculus SCT, except that (1) their profunctors may have 0, 1
or more covariant or contravariant variables but (2) they do not have
a term syntax like our natural transformation syntax and (3) they do
not have functors and restriction of profunctors by functors (since
this structure is not present in a Cartesian
bicategory). Semantically, Cartesian bicategories are somewhat less
general than pro-arrow equipments because in general the set of
functors between two categories can only be recovered from the set of
profunctors if the codomain category is Cauchy complete, in which case
functors are equivalent to adjoint pairs of profunctors.

While we have demonstrated that many common constructions in category
theory can be expressed using the simple combination of homs, tensors
and units, there are several that are out of reach for SCT due to its
restrictive judgmental structure.

First, while we can add products to the calculus, adding in a
\emph{functor category} ${\cat C}^{\cat D}$ would necessitate adding a
syntax for functors of multiple arguments, to allow for the lambda rule:
\[ \inferrule {\alpha:\cat C, \beta:\cat D \vdash e : \cat E} {\alpha:\cat C \vdash \lambda \beta:\cat D. e : \cat E} \]
This would then require also some extension to profunctors for
allowing 0 or more covariant/contravariant dependencies as well.
%
In addition, we could consider extending our semantics with a notion
of \emph{opposite category}, with the property that a covariant
dependence of a profunctor on a variable in a category should be
equivalent to a contravariant dependence on a variable in the opposite
category.
%
This could possibly be done by introducing a modal distinction
$\alpha:^{+}\cat C$ for covariant dependence and $\alpha:^{-}\cat C$
for contravariant dependence, with the opposite category being a
modality that internalizes flipping the variance as a type
constructor.
%
With these two extensions, possibly, the category contexts for
functors and profunctors could be unified as one notion of a sequence
of object variables with variances.
%
With this unification, it would then be simpler to incorporate a third
extension: a universe object, corresponding to an object
$\textrm{Set}$, which would allow us to internalize the profunctor
judgment in the functor judgment, modeling that in classical category
theory, profunctors are the same as certain functors to the category
of Sets.
%
While all of these extensions seem plausible, they all move away to
the simple ordered type theory structure that allowed us to define
units, tensor and hom profunctors so easily. It is unclear how to
define these types in this extended structure. As a simple example, if
profunctors have many variables of mixed variance, rather than exactly
one of each variance, it is unclear what the appropriate
generalization of the transformation context $\Phi$ should be.

>>>>

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
\newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\vett}{VETT}
\newcommand{\ohol}{OHol}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\isSmall}{\,\,\textrm{Small}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}

\newcommand{\smallCats}{\text{SmallCat}}
\newcommand{\Cats}{\text{Cat}}
\newcommand{\varr}[2]{\text{Fun}\,{#1}\,{#2}}
\newcommand{\harr}[2]{\text{Prof}\,{#1}\,{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

\newcommand{\punitinXfromYtoZ}[3]{#2 \mathop{\to_{#1}} #3}
\newcommand{\punitrefl}[1]{\textrm{id}_{#1}}
\newcommand{\punitelimtoYwithkontZ}[2]{\textrm{ind}_{\to}(#1,#2)}
\newcommand{\punitelimkontZatABC}[4]{\textrm{ind}_{\to}(#1,#2, #3, #4)}

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\odot}(#4,#2,#1,#3,#5)}
\newcommand{\tensorelimWfromXtoYkontZ}[4]{\textrm{ind}_{\odot}(#4;#1,#2,#3)}

@@

\begin{figure}
  \begin{mathpar}
    % unit,

    \inferrule
    {\Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}

    \inferrule
    {\Gamma \pipe \alpha:\cat C; d^-\Xi\vdash R \\\\ \Gamma \pipe \alpha : \cat C; d^+\Xi\vdash P}
    {\Gamma \pipe \Xi \vdash \homlallXYtoZ \alpha R P}

  \inferrule
  {\Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \\\\ \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
  {\Gamma \pipe \Xi \vdash \homrallXYtoZ \alpha R P}\\

  \inferrule
    {\Gamma \pipe d^-\Xi \vdash a_1 : \cat C\\\\ \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
    {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2}}

    \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}

    \inferrule{\Gamma\pipe\Xi \vdash R \isaSet \\\\ \Gamma\pipe\Xi \vdash S\isaSet}
              {\Gamma\pipe\Xi \vdash R \times S\isaSet}\\

    \inferrule
    {\Gamma\pipe\beta: \cat D \vdash a : \cat C}
    {\Gamma \pipe\cdot \vdash \punitrefl a : \punitinXfromYtoZ {\cat C} a a}

%%     \inferrule*[right=Unit Elim]
%%     {\Gamma\vdash Q : \harr {\cat C} {\cat C}\and
%%      \Gamma\vdash t : \pendallXdotY {\alpha:\cat C} {\harrapp Q \alpha \alpha}}
%%     {\Gamma \vdash \punitelimtoYwithkontZ Q N :
%%       \pendallXdotY {\alpha_1} {\homrallXYtoZ {\alpha_2} {\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2}} {\harrapp Q {\alpha_1}{\alpha_2}}}}

    \inferrule
    {\Gamma\pipe \alpha:\cat C\vdash t : P[\alpha/\alpha^-;\alpha/\alpha^+]\\\\
     \Gamma\pipe\Phi \vdash s : \punitinXfromYtoZ {\cat C} A B}
    {\Gamma \vdash \punitelimkontZatABC {\alpha. t} A s B : P[A/\alpha^-;B/\alpha^+]}

  \inferrule
  {\Gamma \pipe d^+\Psi_s \vdash b : \cat D\\\\
   \Gamma \pipe \Psi_s \vdash s : P[b/\beta] \\\\
   \Gamma \pipe \Psi_t \vdash t : Q[b/\beta]}
  {\Gamma\pipe\Psi_s \jnctx \Psi_t\vdash \tensorintroatXwithYandZ b s t : \tensorexistsXwithYandZ {\beta : \cat D} P Q}

  \inferrule
  {
   \Gamma\pipe \alpha:\cat C, x:P, \beta:\cat D, y:Q, \gamma:\cat E \vdash t : R\\\\
   \Gamma \pipe\Phi \vdash s : \tensorexistsXwithYandZ {\beta} {P[a/\alpha]}{Q[c/\gamma]}}
  {\Gamma \vdash \tensorelimWfromXtoYkontZ a s b {\alpha.x.\beta.y.\gamma. t} : R[a/\alpha;c/\gamma]}

%%   \inferrule*[right=Tensor-elim]
%%   {\Gamma \pipe \vdash P : \harr {\cat C} {\cat D}\and\Gamma \vdash Q : \harr {\cat D}{\cat E}\and
%%    \Gamma \vdash R : \harr {\cat C}{\cat E}\and
%%    \Gamma \vdash N : \hombinary \alpha {\harrapp P \alpha \beta} \beta {\harrapp Q \beta \gamma} \gamma {\harrapp R \alpha \gamma}}
%%   {\Gamma \vdash \tensorelimfromWXtoYwithkontZ P Q R N : \homunary \alpha {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\gamma} {\harrapp R \alpha \gamma}}
  \inferrule
  {\Gamma \pipe \Phi, x : R, \alpha : \cat C \vdash t : P}
  {\Gamma \pipe \Phi \vdash
   \homrlambdaXatYdotZ {x : R} {\alpha : \cat C} t : \homrallXYtoZ \alpha R P}

  \inferrule
  {\Gamma\pipe\Phi_f \vdash s : \homrallXYtoZ {\alpha:\cat C} R P \\\\
    d^+\Phi_a \vdash a : \cat C\\\\
    \Phi_a \vdash t : R[a/\alpha]
  }
  {\Gamma\pipe \Phi_f \jnctx \Phi_a \vdash \homrappXtoYatZ s t a : P[a/\alpha]}\\

  \inferrule
  {\Gamma\pipe\alpha:\cat C, x : R, \Phi \vdash t : P}
  {\Gamma\pipe\Phi \vdash \homllambdaXatYdotZ {x : R} {\alpha : \cat C} t : \homlallXYtoZ \alpha R P}

  \inferrule
  {\Gamma \pipe d^-\Phi_a \vdash a : \cat C\\\\
   \Gamma \pipe \Phi_a \vdash s : R[a/\alpha] \\\\
   \Gamma \pipe \Phi_f \vdash t : \homlallXYtoZ {\alpha:\cat C} R P}
  {\Gamma \pipe \Phi_a \jnctx \Phi_f \vdash \homlappXtoYatZ t s a : P[a/\alpha]}\\
  % tensor,
  % hom,
  % cartesian products,
  %

  \inferrule{}{\Gamma\pipe \Phi \vdash () :  1}

    \inferrule{\Gamma \pipe\Phi\vdash M : R_1 \times R_2}{\Gamma \pipe\Phi \vdash \pi_i M : R_i}
  \end{mathpar}
  \caption{Set Connectives}
\end{figure}

@@

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma\vdash \smallCats}

    \inferrule
    {}
    {\Gamma\vdash \Cats}

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and\Gamma \vdash \cat D \isaCat}
    {\Gamma\pipe\cdot \vdash \varr{\cat C}{\cat D} \isaSet}

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and\Gamma \vdash \cat D \isaCat}
    {\Gamma\pipe\cdot \vdash \harr{\cat C}{\cat D} \isaSet}

    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash P \isaSet}
    {\Gamma \vdash \pendallXdotY {\alpha:{\cat C}} P \isaTy}

@@

    \inferrule
    {\Gamma \vdash M : \pendallXdotY \alpha P\and
     \Gamma \pipe \beta:\cat D \vdash a : \cat C}
    {\Gamma \pipe \beta:\cat D \vdash \pendappXtoY M a : P[a/\alpha]}
  \end{mathpar}
  \caption{Type Connectives}
\end{figure}

One alternative to proarrow equipments is to axiomatize the bicategory
of categories, profunctors and natural transformations rather than the
virtual double category of categories, functors, profunctors and
natural transformations. An abstract structure along
these lines is a \emph{Cartesian bicategory}\footnote{which is not the
same as a bicategory with products/finite limits, for which a syntax
was developed \citet{fioresaville}} is given in
\citet{carboni-walters,carboni-walters-someonelese}. 
Some preliminary work on a proof system for Cartesian bicateogires was
presented at \citet{jonasfreycartesianbicategories}. This bears strong
similarities to our calculus VETT, except that (1) their profunctors
may have 0, 1 or more covariant or contravariant variables but (2)
they do not have a term syntax like our natural transformation syntax
and (3) they do not have functors and restriction of profunctors by
functors (since this structure is not present in a Cartesian
bicategory). Semantically, Cartesian bicategories are somewhat less
general than virtual equipments because in general the set of
functors between two categories can only be recovered from the set of
profunctors if the codomain category is Cauchy complete, in which case
functors are equivalent to adjoint pairs of profunctors.

While we have demonstrated that some common constructions in category
theory can be expressed using the simple combination of homs, tensors
and units, there are several that are out of reach for SCT due to its
restrictive judgmental structure. These require extensions to the
judgmental structure that would move beyond virtual equipments.

First, while we can add products to the calculus, adding in a
\emph{functor category} ${\cat C}^{\cat D}$ would necessitate adding a
syntax for functors of multiple arguments, to allow for the lambda rule:
\[ \inferrule {\alpha:\cat C, \beta:\cat D \vdash e : \cat E} {\alpha:\cat C \vdash \lambda \beta:\cat D. e : \cat E} \]
This would then require also some extension to profunctors for
allowing 0 or more covariant/contravariant dependencies as well.
%
Semantically, this would correspond to generalizing from virtual
equipments to some kind of (possibly cartesian) \emph{monoidal closed}
virtual equipments.

In addition, we could consider extending our semantics with a notion
of \emph{opposite category}, with the property that a covariant
dependence of a profunctor on a variable in a category should be
equivalent to a contravariant dependence on a variable in the opposite
category.
%
This could possibly be done by introducing a modal distinction
$\alpha:^{+}\cat C$ for covariant dependence and $\alpha:^{-}\cat C$
for contravariant dependence, with the opposite category being a
modality that internalizes flipping the variance as a type
constructor.
%
With these two extensions, possibly, the category contexts for
functors and profunctors could be unified as one notion of a sequence
of object variables with variances.
%
With this unification, it would then be simpler to incorporate a third
extension: a universe object, corresponding to an object
$\textrm{Set}$, which would allow us to internalize the profunctor
judgment in the functor judgment, modeling that in classical category
theory, profunctors are the same as certain functors to the category
of Sets.
%
While all of these extensions seem plausible, they all move away to
the simple ordered type theory structure that allowed us to define
units, tensor and hom profunctors so easily. It is unclear how to
define these types in this extended structure. As a simple example, if
profunctors have many variables of mixed variance, rather than exactly
one of each variance, it is unclear what the appropriate
generalization of the transformation context $\Phi$ should be.

@@

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma\vdash \smallCats}

    \inferrule
    {}
    {\Gamma\vdash \Cats}

    \inferrule
    {\Gamma \vdash \cat C : \Sort \and\Gamma \vdash \cat D : \Sort}
    {\Gamma\pipe\cdot \vdash \varr{\cat C}{\cat D} \isaSet}

@@

@@

    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash P \isaSet}
    {\Gamma \vdash \pendallXdotY {\alpha:{\cat C}} P \isaTy}

    \inferrule
    {\Gamma \vdash M : \pendallXdotY \alpha P\and
     \Gamma \pipe \beta:\cat D \vdash a : \cat C}
    {\Gamma \pipe \beta:\cat D \vdash \pendappXtoY M a : P[a/\alpha]}
  \end{mathpar}
  \caption{Type Connectives}
\end{figure}
\begin{figure}
  \begin{mathpar}
    \inferrule
    {\Gamma \pipe \alpha:\cat C\vdash P \isaSet}
    {\Gamma \vdash \pendallXdotY {\alpha:{\cat C}} P \isaTy}

    \inferrule*[right=2Cell$\eta$]
    {\Gamma \vdash M : \pendallXdotY \alpha P }
    {\Gamma \vdash M = \pendlambdaXdotY \alpha {\pendappXtoY M \alpha} : \pendallXdotY \alpha P}
  \end{mathpar}
  \caption{End}
  \label{fig:end}
\end{figure}
\section{Semantics}

@@
